---
title: "Causal Inference Workshop"
output:
  html_document
---

# Introduction

Welcome to the **Causal Inference Workshop**! This notebook explores fundamental causal inference concepts, focusing on confounding, selection bias, and essential causal inference methods.

**Essential Question**: Under what conditions can we make causal inferences using observational data?

------------------------------------------------------------------------

Our goals for today's workshop are:

1.  Understand the conceptual framework of causal inference, including:
    -   **Potential Outcomes Framework**
    -   **Stable Unit Treatment Value Assumption (SUTVA)**
    -   **Assignment Mechanism**
2.  Recognize observational data limitations and strategies to address them
3.  Apply causal inference methods like propensity score matching (PSM) and instrumental variables (IV) across datasets.
4.  Implement real-world applications with existing observational datasets

------------------------------------------------------------------------

# 📊 Dataset Descriptions

## 1️⃣ Lalonde Dataset

**Context**: Evaluates the National Supported Work (NSW) Demonstration, a job training program.

-   **Treatment**: Participation in the job training program.
-   **Outcome**: Post-program earnings (`re78`).
-   **Covariates**: Age, education, race, pre-program earnings (`re74`, `re75`).

## 2️⃣ NHANES Dataset

**Context**: NHANES assesses health and nutrition. We explore physical activity's effect on blood pressure.

-   **Treatment**: Regular physical activity (`PhysActive`).
-   **Outcome**: Systolic blood pressure (`BPXSY1`).
-   **Covariates**: Age, income, and BMI.

## 3️⃣ PSID Dataset

**Context**: PSID follows U.S. families over time, often used to study intergenerational effects.

-   **Treatment**: Parental education level (`parent_education`).
-   **Outcome**: Child education level (`child_education`).
-   **Covariates**: Household income and neighborhood quality.

------------------------------------------------------------------------

## Key Concept 1: Potential Outcomes Framework

-   **The Potential Outcomes Framework** defines causal effects by imagining possible outcomes for each subject under different conditions.

-   **Counterfactual Reasoning**: Essentially 'what if' scenarios. Only one outcome is observed. The difference **Y(1) - Y(0)** is the individual treatment effect.

### 🥊 Challenge 1: Potential Outcomes with Lalonde Dataset

```{r}
# Potential outcomes example with hypothetical Lalonde data
hypothetical_data <- data.frame(
  Individual = 1:3,
  Y1 = c(5, 6, 8),  # Income if treated
  Y0 = c(3, 6, 4)   # Income if not treated
)

# Calculate Individual Treatment Effects (ITE)
hypothetical_data$ITE <- hypothetical_data$Y1 - hypothetical_data$Y0
# Calculate Average Treatment Effect (ATE)
ATE <- mean(hypothetical_data$ITE)
hypothetical_data
ATE

```

In causal inference language, we refer to the measures ITE and ATE that we calculated above as the *causal estimands*

**Causal Estimand** refers to the different between two potential outcomes, that is the outcome where a subject receives the treatment Y(1) and the subject that does not Y(1). In the exercise above, we calculate the Individual Treatment Effect (ITE) and the Average Treatment Effect (ATE) for a population.

# Confounding and Selection Bias

**Confounding** arises when a variable influences both the treatment and outcome, creating bias in estimating causal effects.

**Selection bias** occurs when the treatment and control groups are not comparable due to systematic differences (e.g., motivation, socioeconomic status).

### 🥊 Challenge 2: Identifying confounders in the Lalond Dataset

-   Use the Lalonde dataset to show differences in pre-treatment characteristics between treated and untreated groups
-   Plot these differences using dplyr

💡 **Tip**: Look up the Lalonde dataset to find more information about the variables in the dataset and what information the survey holds. Find a variable (or variables) that show issues of confounding or selection bias. Explain why and plot your results showing treated vs. untreated group.

```{r}
library(dplyr)
# Education level by treatment status
ggplot(lalonde, aes(x = factor(treat), y = age)) +
  geom_boxplot() +
  labs(title = "Distribution of Age by Treatment Status", x = "Treatment Status", y = "Years of Age")


```

Answer: clear difference because age might be a confounder as it influences both the likelihood of participating in the job training program AND the outcome (post-program earnings)

Implications for selection bias:

The systematic difference in age between treated and control groups suggest selection bias, as certain age groups appear more likely to participate in the program.

------------------------------------------------------------------------

## Key Concept 2: SUTVA (Stable Unit Treatment Value Assumption)

SUTVA requires **no interference** between units and **no hidden variations** in treatment.

-   **No interference**: An individual's treatment does not affect another's outcome.
-   **No hidden variation**: The treatment is consistent across all individuals.

### 🥊 Challenge 3: Violations of SUTVA

a.  Identify potential violations of SUTVA. Example scenarios:

-   **Vaccination**: Immunity provides indirect benefits to others.

```{r}

```

-   **Job Training**: Affects job availability, impacting others.

```{r}

```

b.  Take a look at the following scenario:

-   Each individual is assigned to either participate in a job training program (treatment) or not (control).

-   The outcome we’re interested in is job readiness or income after the program.

Consider the following:

-   Treatment Status: Individuals are randomly assigned to either receive training (Treatment = 1) or not (Treatment = 0).

-   Neighbor Influence: Each individual’s outcome could also be influenced if their "neighbor" (e.g., a friend or family member) received training, modeled by neighbors_treated.

-   Outcome (Job Readiness Score):The outcome is calculated based on:

    -   A base score of 10.

    -   An increase of 3 points if the individual received training.

    -   An additional 1.5 points if their neighbor received training.

```{r}
# Set seed for reproducibility
set.seed(123)

# Simulate 10 individuals in a community
n <- 10
treatment <- sample(c(0, 1), n, replace = TRUE)  # Randomly assign job training (1) or no training (0)

# Assume each individual has a "friend" or "neighbor" who could be influenced by their training status
# Here, the influence is simplified to mean the previous individual in the sequence
neighbors_treated <- c(0, treatment[-n])  # Count if the previous person was in treatment

# Outcome depends on whether the individual received training and if their "neighbor" received training
# Outcome: post-program job readiness score (hypothetical)
outcome <- 10 + 3 * treatment + 1.5 * neighbors_treated + rnorm(n, mean = 0, sd = 2)

# Show results
data <- data.frame(
  Individual = 1:n,
  Treatment_Status = ifelse(treatment == 1, "Received Training", "No Training"),
  Neighbor_Treatment_Status = ifelse(neighbors_treated == 1, "Neighbor Trained", "Neighbor Not Trained"),
  Job_Readiness_Score = round(outcome, 2)
)

# Display the simulated data
print(data)

```

Does this illustrate a scenario where SUTVA does not hold? Explain.

```{r}
##answer here 

#nterference occurs because participants might share valuable skills or job-seeking tips with their friends who are not in the program, which could improve the job outcomes for the control group.

#No Interference is violated because one individual's treatment (job training) can affect their neighbor's job readiness score.

#No Hidden Variations in Treatment could also be discussed if different individuals receive varying levels of support within the program, though this is not explicitly simulated here.

```

------------------------------------------------------------------------

## Key Concept 3: Assignment Mechanism

In observational studies, **assignment mechanisms** may introduce bias due to self-selection or confounding factors.

The **assignment mechanism** determines who receives the treatment and who does not. It is crucial in causal inference because it affects the comparability of treatment and control groups.

Types of assignment mechanisms:

-   **Random Assignment:** Treatment is assigned randomly, helping to create comparable groups. 

-   **Quasi-Experimental:** Assignment is not random but may follow specfic rules (e.g., eligibility based on age or income, the draft)

-   **Self-selection**: Individuals decide whether to participate, often leading to selection bias. 

### 🥊 Challenge 4: Simulating Random Assignment

```{r}
# Example random assignment simulation
set.seed(123)
n <- 20  # Number of participants
assignment <- sample(c('Treatment', 'Control'), n, replace = TRUE)
table(assignment)
```

------------------------------------------------------------------------

# 🎬 Demo: Propensity Score Matching (PSM) to deal with issues of selection 


Propensity Score Matching (PSM) is used to control for confounding variables by balancing covariates across treatment and control groups.

### 🥊 Challenge 5: Applying PSM

```{r}
library(MatchIt)
# Apply PSM on the Lalonde dataset
psm_model <- matchit(treat ~ age + educ + race + re74 + re75, data = lalonde, method = 'nearest')
summary(psm_model)
plot(psm_model)
```
